---
title: "Introduction to metafrontier"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to metafrontier}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## What is a metafrontier?

In efficiency analysis, we often study firms that operate under
fundamentally different technologies. Steel producers using electric arc
furnaces (EAF) face a different production possibility set than those using
the blast furnace-basic oxygen furnace (BF-BOF) route. Hospitals in rural
areas face different constraints than urban ones. Banks in developing
economies operate under different regulatory and technological environments
than those in advanced economies.

Standard stochastic frontier analysis (SFA) or data envelopment analysis
(DEA) applied to the pooled sample implicitly assumes all firms share the
same technology -- an assumption that may be unrealistic. Estimating
separate frontiers for each group solves this problem but makes efficiency
scores incomparable across groups: a firm that is 90\% efficient relative
to a less advanced group frontier may actually be less productive than a
firm that is 70\% efficient relative to a more advanced frontier.

The **metafrontier** framework, introduced by Battese, Rao, and O'Donnell
(2004) and extended by Huang, Huang, and Liu (2014) and O'Donnell, Rao,
and Battese (2008), resolves this by:

1. Estimating **group-specific frontiers** for each technology group
2. Estimating a **metafrontier** that envelops all group frontiers
3. Decomposing efficiency into two components:

$$TE^*_i = TE_i \times TGR_i$$

where:

- $TE_i$ is efficiency relative to the **group frontier** (within-group
  inefficiency)
- $TGR_i$ is the **technology gap ratio**, measuring how close the group
  frontier is to the metafrontier (between-group technology gap)
- $TE^*_i$ is efficiency relative to the **metafrontier** (overall
  efficiency)

The `metafrontier` package provides a unified interface for estimating
metafrontier models using both SFA and DEA approaches.


## Quick start

```{r setup}
library(metafrontier)
```

### Simulate data

The package includes `simulate_metafrontier()` for generating data from a
known data-generating process. This is useful for Monte Carlo studies and
for learning the package.

```{r simulate}
sim <- simulate_metafrontier(
  n_groups = 3,
  n_per_group = 200,
  beta_meta = c(1.0, 0.5, 0.3),  # intercept, elasticity_1, elasticity_2
  tech_gap = c(0, 0.25, 0.5),    # intercept shifts (0 = best technology)
  sigma_u = c(0.2, 0.3, 0.4),    # inefficiency SD per group
  sigma_v = 0.15,                 # noise SD
  seed = 42
)

str(sim$data[, c("log_y", "log_x1", "log_x2", "group")])
table(sim$data$group)
```

The simulation generates a Cobb-Douglas frontier:
$$\ln y_i = \beta_0^{(j)} + \beta_1 \ln x_{1i} + \beta_2 \ln x_{2i} + v_i - u_i$$
where the intercept $\beta_0^{(j)} = \beta_0^* - \delta_j$ is shifted down
from the metafrontier by the technology gap $\delta_j$ for group $j$.

### Estimate the metafrontier

```{r estimate}
fit <- metafrontier(
  log_y ~ log_x1 + log_x2,
  data = sim$data,
  group = "group",
  method = "sfa",
  meta_type = "deterministic"
)

fit
```


## Deterministic SFA metafrontier (Battese, Rao, and O'Donnell, 2004)

The deterministic metafrontier is estimated in two stages:

1. **Stage 1**: Fit separate SFA models for each group via maximum
   likelihood.
2. **Stage 2**: Find metafrontier coefficients $\hat\beta^*$ by minimising
   $$\sum_i \left[\ln f(x_i; \hat\beta^*) - \ln f(x_i; \hat\beta_j)\right]^2$$
   subject to the constraint that the metafrontier envelops all group
   frontiers: $\ln f(x_i; \hat\beta^*) \ge \ln f(x_i; \hat\beta_j)$ for
   all $i$ and $j$.

This is the default method:

```{r deterministic}
fit_det <- metafrontier(
  log_y ~ log_x1 + log_x2,
  data = sim$data,
  group = "group",
  meta_type = "deterministic"
)

summary(fit_det)
```


## Stochastic metafrontier (Huang, Huang, and Liu, 2014)

The stochastic metafrontier replaces the LP in Stage 2 with a second-stage
SFA, using the fitted group frontier values as the dependent variable:

$$\ln \hat{f}(x_i; \hat\beta_j) = x_i'\beta^* + v^*_i - u^*_i$$

where $u^*_i \ge 0$ captures the technology gap stochastically. This
provides a distributional framework for the TGR, enabling standard errors
and hypothesis testing.

```{r stochastic}
fit_sto <- metafrontier(
  log_y ~ log_x1 + log_x2,
  data = sim$data,
  group = "group",
  meta_type = "stochastic"
)

summary(fit_sto)
```

The stochastic metafrontier provides a variance-covariance matrix:

```{r vcov}
vcov(fit_sto)
```


## DEA-based metafrontier

For a nonparametric approach, set `method = "dea"`:

```{r dea}
fit_dea <- metafrontier(
  log_y ~ log_x1 + log_x2,
  data = sim$data,
  group = "group",
  method = "dea",
  rts = "vrs"
)

fit_dea
```

The DEA metafrontier computes:

1. Group-specific DEA efficiencies under variable returns to scale (VRS),
   constant returns to scale (CRS), or other assumptions.
2. A pooled DEA across all observations to form the metafrontier.
3. $TGR_i = TE^{pool}_i / TE^{group}_i$.


## Extracting results

### Efficiency scores

Use `efficiencies()` to extract the three components of the decomposition:

```{r efficiencies}
te <- efficiencies(fit_det, type = "group")
tgr <- efficiencies(fit_det, type = "tgr")
te_star <- efficiencies(fit_det, type = "meta")

# Verify the fundamental identity: TE* = TE x TGR
all.equal(te_star, te * tgr)
```

### Technology gap ratio

The `technology_gap_ratio()` function returns TGR values grouped by
technology:

```{r tgr}
tgr_by_group <- technology_gap_ratio(fit_det)
lapply(tgr_by_group, summary)
```

For a formatted summary table:

```{r tgr-summary}
tgr_summary(fit_det)
```

### Coefficients

```{r coefs}
# Metafrontier coefficients
coef(fit_det, which = "meta")

# Group-specific coefficients
coef(fit_det, which = "group")
```

### Model information

```{r model-info}
# Log-likelihood (sum of group log-likelihoods for deterministic)
logLik(fit_det)

# Number of observations
nobs(fit_det)

# AIC and BIC (available automatically via logLik method)
AIC(fit_det)
```


## Visualisation

The package provides four built-in plot types:

### TGR distributions

```{r plot-tgr, fig.height=4}
plot(fit_det, which = "tgr")
```

### Efficiency scatter

```{r plot-eff, fig.height=4}
plot(fit_det, which = "efficiency")
```

Points below the 45-degree line indicate a technology gap (TE* < TE). The
vertical distance from the line reflects the TGR.

### Efficiency decomposition

```{r plot-decomp, fig.height=4, fig.width=9}
plot(fit_det, which = "decomposition")
```

Side-by-side boxplots of TE, TGR, and TE* by group.


## Hypothesis testing

### Poolability test

The poolability test evaluates whether group-specific frontiers are
statistically different from a single pooled frontier:

```{r poolability}
poolability_test(fit_det)
```

A significant result (small p-value) indicates that the technology groups
have genuinely different production technologies, justifying the
metafrontier approach.


## Inefficiency distributions

The package supports three distributional assumptions for the one-sided
inefficiency term $u_i$ in SFA:

```{r distributions, eval=FALSE}
# Half-normal (default): u ~ |N(0, sigma_u^2)|
fit_hn <- metafrontier(log_y ~ log_x1 + log_x2,
                       data = sim$data, group = "group",
                       dist = "hnormal")

# Truncated normal: u ~ N+(mu, sigma_u^2)
fit_tn <- metafrontier(log_y ~ log_x1 + log_x2,
                       data = sim$data, group = "group",
                       dist = "tnormal")

# Exponential: u ~ Exp(1/sigma_u)
fit_exp <- metafrontier(log_y ~ log_x1 + log_x2,
                        data = sim$data, group = "group",
                        dist = "exponential")
```


## Comparing true and estimated values

Since we used simulated data, we can compare estimated values against the
truth:

```{r compare-truth}
# True vs estimated metafrontier coefficients
cbind(
  True = sim$params$beta_meta,
  Estimated = coef(fit_det, which = "meta")
)

# True vs estimated mean TGR by group
true_tgr <- tapply(sim$data$true_tgr, sim$data$group, mean)
est_tgr <- tapply(fit_det$tgr, fit_det$group_vec, mean)
cbind(True = true_tgr, Estimated = est_tgr)

# Correlation between true and estimated efficiency
cor(sim$data$true_te, fit_det$te_group)
cor(sim$data$true_te_star, fit_det$te_meta)
```


## References

- Battese, G.E., Rao, D.S.P. and O'Donnell, C.J. (2004). A metafrontier
  production function for estimation of technical efficiencies and
  technology gaps for firms operating under different technologies.
  *Journal of Productivity Analysis*, 21(1), 91--103.

- Huang, C.J., Huang, T.-H. and Liu, N.-H. (2014). A new approach to
  estimating the metafrontier production function based on a stochastic
  frontier framework. *Journal of Productivity Analysis*, 42(3), 241--254.

- O'Donnell, C.J., Rao, D.S.P. and Battese, G.E. (2008). Metafrontier
  frameworks for the study of firm-level efficiencies and technology
  ratios. *Empirical Economics*, 34(2), 231--255.
